Fast gravitational wave parameter estimation without compromises

### 2023-01-23 src/tex/ms.tex

To compute the integral shown in Eq.~\eqref{eq: innerproduct}, we need to
evaluate a chosen waveform model $h(\mathbf{\theta})$ at a number of frequency
sample points. This makes evaluating the likelihood function often the most
computationally intensive part of PE. The most accurate waveform model is
numerical relativity (NR), which obtains waveforms by directly solving the
Einstein equations numerically for a given system. However, depending on the
source parameters, generating one time series of strain can take a day to half
a year, which makes NR prohibitively expensive for PE. To circumvent this
problem, there are several families of waveform ``approximants'', including the
\textsc{IMRPhenom} family \cite{Khan:2015jqa, Garcia-Quiros:2020qpx}, the
\textsc{SEOB} family \cite{PhysRevD.89.061502}, and the NR surrogate
family\cite{Varma:2019csw}. For shorter events, such as a $30-30\ M_{\odot}$
binary black hole, one waveform call could take $10\text{ms}$ to ${\sim}1s$
\kw{verify number in lalsuite}. For longer events, such as a $1.4-1.4\
M_{\odot}$ binary neutron star, the evaluation time could go up to \kw{Fill}.
Since one needs to evaluate the likelihood millions of times during
sampling,\footnote{A typical PE run with \texttt{Bilby} takes ${>}10^6$
likelihood evaluations to converge.} the computational cost in evaluating the
waveform accumulates and is the main reason of the long runtime of GW PE.

@@

To obtain a reference waveform, we currently use the \texttt{differential
evolution} algorithm \cite{Storn1997DifferentialE} available in the
\textsc{scipy} package \cite{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation.

@@

Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. We use the package
\texttt{flowMC} \cite{2022arXiv221106397W}. The pseudocode of the algorithm is
given in Algorithm \ref{alg:cap}.

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\texttt{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 5 minutes \kw{Benchmark it
in the code and make this number precise.} for GW170817. There are two ways to
reduce this time.

>>>>

To compute the integral shown in Eq.~\eqref{eq: innerproduct}, we need to
evaluate a chosen waveform model $h(\mathbf{\theta})$ at a number of frequency
sample points. This makes evaluating the likelihood function often the most
computationally intensive part of PE. The most accurate waveform model is
numerical relativity (NR), which obtains waveforms by directly solving the
Einstein equations numerically for a given system. However, depending on the
source parameters, generating one time series of strain can take a day to half
a year, which makes NR prohibitively expensive for PE. To circumvent this
problem, there are several families of waveform ``approximants'', including the
\textsc{IMRPhenom} family \cite{Khan:2015jqa, Garcia-Quiros:2020qpx}, the
\textsc{SEOB} family \cite{PhysRevD.89.061502}, and the NR surrogate
family\cite{Varma:2019csw}. For shorter events, such as a $30-30\ M_{\odot}$
binary black hole, one waveform call could take $10\text{ms}$ to ${\sim}1s$
\kw{verify number in lalsuite}. For longer events, such as a $1.4-1.4\
M_{\odot}$ binary neutron star, the evaluation time could go up to \kw{Fill}.
Since one needs to evaluate the likelihood millions of times during
sampling,\footnote{A typical PE run with \textsc{Bilby} takes ${>}10^6$
likelihood evaluations to converge.} the computational cost in evaluating the
waveform accumulates and is the main reason of the long runtime of GW PE.

@@

To obtain a reference waveform, we currently use the \textsc{differential
evolution} algorithm \cite{Storn1997DifferentialE} available in the
\textsc{scipy} package \cite{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation.

@@

Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. We use the package
\textsc{flowMC} \cite{2022arXiv221106397W}. The pseudocode of the algorithm is
given in Algorithm \ref{alg:cap}.

@@

% Faster maximum likelihood finder
Besides compilation, there is in principle also overhead from finding the
reference waveform used in for heterodyning the likelihood. Since the
\textsc{differential evolution} algorithm we currently use has not been
implemented in \textsc{Jax}, and the \textsc{Jax} waveform we use is not
compatible with the parallelization scheme in the \textsc{scipy} library,
maximizing the likelihood currently takes us around 5 minutes \kw{Benchmark it
in the code and make this number precise.} for GW170817. There are two ways to
reduce this time.
